---
title: "R Codes - graphical descriptive statistics"
author: "Subham Kailthya"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

## Load libraries

```{r}
library("knitr")          # knitting doc
library("dplyr")          # data wrangling
library("ggplot2")        # plotting data
library("janitor")        # manage tables
library("sf")             # spatial data
library("rnaturalearth")  # spatial database
```

## Import Freedom House Data

Here we import FH data from `/path/link/<data-file>`, drop the variable `Region_Code`, convert `Status`, `Region_Name` and `is_ldc` to factor data type, and keep observations corresponding to the year 2020.

`select()`, `mutate()` and `filter()` are functions from the `ddplyr` package. To know more about a function, say `mutate()`, type `help("mutate")` or `?mutate()` in `R` console.

```{r}
path <- "https://raw.githubusercontent.com/subhamk/"
link <- "ec122_share/main/data/"

freedom <- readr::read_csv(paste0(path, link, "freedom.csv")) # import csv

freedom_2020 <- freedom %>%
  dplyr::select(-Region_Code) %>%           # drop Region_Code
  mutate(Status = factor(Status),
         Region_Name = factor(Region_Name),
         is_ldc = factor(is_ldc)) %>%       # convert to factor
  filter(year == 2020)                      # keep data for the year 2020

head(freedom_2020)                          # print first few lines of the data
```


The function `glimpse()` provides a quick overview of the dataset.
```{r, echo=TRUE}
glimpse(freedom_2020)
```

## Cross Tables

This tabulates the number of observations per cell where each cell is a combination of `Stutus` and `is_ldc`.

```{r crosstabs}
freedom_2020 %>% 
  tabyl(is_ldc, Status) %>%             # create table
  adorn_totals(c("row", "col")) %>%     # add row and col totals
  kable()                               # print table
```

## Bar Chart

The following code chunk plots a bar chart of average `CL` score by group `Status`. `summarise` creates a new data (`tibble` in tidy language) which contains the mean of `CL` and we name this `fh_ranking`.

```{r barchart-data}
fh_ranking <- freedom_2020 %>%          # use 2020 FH data
  group_by(Status) %>%                  # group data by Status
  summarise(mean_cl = mean(CL, na.rm = TRUE))   # calculate the mean by group var.
fh_ranking
```

This reproduces the bar chart in the slides. We use `ggplot2` library to produce charts. Hadley Wickham's [ggplot2 book](ggplot2: Elegant Graphics for Data Analysis (3e)) is a detailed resource.
```{r barchart-plot}
ggplot(data = fh_ranking, aes(x = Status, y = mean_cl)) +
  geom_bar(stat = 'identity',
           fill = 'steelblue',
           alpha = 0.6) +
  geom_text(
    aes(label = round(mean_cl, 2)),
    colour = "black",
    size = 3,
    vjust = 1.5,
    position = position_dodge(.9)
  ) +
  labs(
    title = "Civil Liberty Rating by Status",
    subtitle = "Note: Smaller CL values indicate fewer restrictions",
    x = "Status",
    y = "CL Rating"
  ) +
  theme_bw()
```

## Bar Chart with Multiple Groups

Here, we calculate the mean by both `Status` and `is_ldc` and present data for each of the groups side by side. We construct a new variable `label_y` -- the cumulative sum of `mean_cl` -- which helps in placing the value labels in the stacked bar plot (next section) in the right place.

```{r multibar-data}
fh_ranking_m <- freedom_2020 %>%
  group_by(Status, is_ldc) %>%          # group data
  summarise(mean_cl = mean(CL, na.rm = TRUE)) %>%   # get averages
  mutate(label_y = cumsum(mean_cl))     # cumulative sum
fh_ranking_m
```


This reproduces multiple bar plot in the slides.
```{r multibar-plot}
ggplot(data = fh_ranking_m, aes(x = Status, y = mean_cl, fill = is_ldc)) +
  geom_bar(stat = 'identity', position = position_dodge()) +
  scale_fill_brewer(palette = "Blues") +
  geom_text(
    aes(label = round(mean_cl, 2)),
    colour = "black",
    size = 3,
    vjust = 1.5,
    position = position_dodge(.9)
  ) +
  labs(
    title = "Civil Liberty Rating by LDC and Status",
    subtitle = "Note: Smaller values indicate fewer constraints",
    x = "Status",
    y = "CL Rating"
  ) +
  theme_bw()
```

## Stacked Bar Chart

We stack the bars instead of presenting them side by side. This is particularly useful when there are several categories for a variable of interest.
```{r stacked-barplot}
ggplot(data = fh_ranking_m, aes(x = Status, y = mean_cl, fill = is_ldc)) +
  geom_bar(stat = 'identity') +
  scale_fill_brewer(palette = "Blues") +
  geom_text(
    aes(y = label_y, label = round(mean_cl, 2)),
    colour = "black",
    size = 3,
    vjust = 1.5
  ) +
  labs(
    title = "Civil Liberty Rating by LDC and Status",
    subtitle = "Note: Smaller value indicates fewer constraints",
    x = "Status",
    y = "CL Rating"
  ) +
  theme_bw()
```

## Stacked Bar Chart - Proportional

We could stack proportions instead.
```{r stackedprop}
ggplot(data = fh_ranking_m, aes(x = Status, y = mean_cl, fill = is_ldc)) +
  geom_col(position = "fill") +
  scale_fill_brewer(palette = "Blues") +
  labs(
    title = "Civil Liberty Rating by LDC and Status",
    subtitle = "Note: Smaller value indicates fewer constraints",
    x = "Status",
    y = "CL Rating"
  ) +
  theme_bw()
```

## Pie Chart

```{r pie-data}
dat <- freedom_2020 %>%
  janitor::tabyl(Status) %>%    # creates a table
  arrange(desc(n))              # descending order 
dat
```

```{r pie, fig.align='center', out.width="60%"}
ggplot(dat, aes(x = "", y = percent, fill = Status)) +
  geom_bar(stat = "identity",
           width = 1,
           alpha = 0.6) +
  coord_polar("y", start = 0) +
  theme_void()
```

## Merging data

First, we load PWT 10.1 and name this data object `pwt`. We then create a new data object `pwt_1` which subsets `pwt` to keep only GDP data for the year 2010.  

```{r, echo=TRUE}
pwt <- readr::read_csv(paste0(path, link, "pwt1001.csv"))
pwt_1 <- pwt %>% 
  filter(year == 2010) %>% 
  dplyr::select(countrycode, country, rgdpo)

head(pwt_1)
```

We then merge FH data with GDP data based on `country` (country names). We drop observations where `rgdpo` are missing (indicated by `NA` in `R`) and then take the log of GDP (`log_rgdpo`).
```{r, echo=TRUE}
freedom_gdp <- freedom_2020 %>% 
  dplyr::left_join(pwt_1, by = join_by(country)) %>% 
  tidyr::drop_na(rgdpo) %>% 
  mutate(log_rgdpo = log(rgdpo))
head(freedom_gdp)
```

## Histogram

The code plots the distribution of log GDP for the year 2010 across all the countries in our sample. We specify the breaks in `br`.
```{r hist, echo=TRUE}
br <- c(7, 9, 10, 12, 14, 16, 18)        # specify breaks for histogram

ggplot(data = freedom_gdp, aes(x = log_rgdpo, stat(density))) +
  geom_histogram(breaks = br,
                 fill = 'steelblue',
                 alpha = 0.6) +
  ylim(0, 0.2) +
  labs(
    title = 'Histogram of log GDP',
    subtitle = 'User supplied bin width',
    x = 'log GDP',
    y = 'RFD'
  ) +
  theme_bw()
```

## Histogram - Automatic bin selection

Here, we plot the distribution of `log_rgdpo` but allow `R` to automatically select bin size. This results in a more elegant plot.

```{r, echo=TRUE}
ggplot(data = freedom_gdp, aes(x = log_rgdpo, y = stat(density))) +
  geom_histogram(fill = 'steelblue', alpha = 0.6) +
  labs(
    title = 'Distribution of Real GDP in PPP',
    subtitle = 'Data source: PWT 10.01',
    x = 'log GDP',
    y = 'Count'
  ) +
  theme_bw()
```


## Time Series Plot

We might be interested in how a variable evolves over time. Suppose we are interested in how the UK economy has grown over time.

To do this, we start with `pwt` data object, subset observations corresponding to the UK, select the variables `year` and `rgdpo` that are of interest. Note that `rgdpo` is in chained PPP. We then create a new variable `log_gdp` which log transforms real UK GDP.

```{r}
uk_gdp <- pwt %>% 
  filter(countrycode == "GBR") %>% 
  dplyr::select(year, rgdpo) %>% 
  mutate(log_gdp = log(rgdpo))
head(uk_gdp)
```


We then plot `log_gdp` against `year` to show how the UK economy has grown over time.
```{r tsplot}
ggplot(data = uk_gdp, aes(x = year, y = log_gdp)) +
  geom_line(linewidth = 2, col = 'steelblue') +
  labs(title = 'UK GDP chained PPP',
       x = 'Year',
       y = "GDP (in logs)") +
  theme_bw()
```


## Geo-spatial data

`R` has advanced features to deal with spatial data. We obtain the map of the world from naturalearth, merge `fh_ranking_2020` to get the civil liberty scores by country and plot it on a map.

```{r}
world <- ne_countries(scale = "medium", returnclass = "sf")

fh_ranking_2020 <- freedom_2020 %>%
  mutate(
    country = replace(
      country,
      country == "United Kingdom of Great Britain and Northern Ireland",
      "United Kingdom"
    )
  ) %>%
  dplyr::select(CL, PR, country)

world <- world %>%
  left_join(fh_ranking_2020, by = join_by(geounit == country))
```

```{r}
ggplot(data = world) +
  geom_sf(aes(fill = factor(CL))) +
  labs(title = 'Civil Liberty Ranking in 2020',
       subtitle = 'Smaller score indicates fewer constraints') +
  theme_bw()
```
